---
title: 'Lecture 4: Efficient coding practices'
author: "Prof. Ariel Ortiz-Bobea"
date: "AEM 6850"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Plan for today

1.  Motivation
2.  My philosophy
3.  Example from a recent study
4.  Organizing projects/scripts/code
5.  Guiding principles
6.  Getting started with your homework

# Disclaimers

-   I'm not a computer scientist
-   The ideas here reflect my experience and that of others
-   Although our workshop is in R, many of these ideas apply to other languages

# Motivation

Good code...

-   Increases confidence in your work
-   Increases the chances other researchers will reproduce, replicate and cite your work
-   Saves you (and others) time
-   Reduces the chances of making mistakes
-   Is easy to understand
-   May be portable

# Coding philosophy

You are writing code for OTHERS (that includes yourself in $t+1$)

# Organizing a project

-   A project comes with general instructions

    -   It could be a readme.txt file
    -   It should specify the authors, what the project is about and describe files

-   Folders for different types of files

    -   Don't mix raw data with "cleaned" data

-   No single best way of organization but this is a good start:

    -   Folder for code (no subfolders)

    -   Folder(s) for input(s)

        -   A folder for raw data (which you never modify)
        -   A folder for cleaned/modified data (which you create in the script files)

    -   Folder(s) for output(s)

        -   A folder for images / tables / etc.

    -   Folder for packages (optional)

# Example of folder structure

-   code (all script files including RStudio project)

-   data

    -   raw (data as downloaded from web, can be large)
    -   clean (data needed for analysis)

-   output (that show up in the paper only)

    -   figures
    -   tables

-   paper

-   presentations

-   Others: meeting notes, documentation, etc.

# Organizing your script files

-   A script file should accomplish a specific and well-identified task in a project

    -   Example of tasks: cleaning data, regression analysis, etc.

-   Never have all your code in a single script file (unless it is a tiny project. e.g. your homework)

-   Script files should have intuitive names and follow a work flow:

    -   1.1_clean_economic_data.R
    -   1.2_clean_pollution_data.R
    -   1.3_prepare_regression_data.R
    -   2.1_regression_analysis.R
    -   2.2_post_estimation_analysis.R

# Good practice to create special script files

-   A master script file (e.g. master.R) that runs all project scripts file in the correct order

    -   Installs/loads all necessary packages
    -   Simplify life for reproducers
    -   Typically created at the very end of the project

-   A file containing user-defined functions (e.g. functions.R)

    -   improves automation if used in several scripts
    -   invites you to create more portable functions for future projects


# Code within a script file

-   A script file should have a clear structure with sections

-   Include a title and description of what the script file does

-   There should be a "preliminary section" where you set things up

    -   Import data
    -   Load necessary packages
    -   Specify any settings or parameters for the analysis
    -   Maybe define functions that are only used in this script file, etc.

-   This section is followed by one or more sections where tasks are performed

    -   No "preliminary" stuff here (e.g. importing data)


# Guiding Principles 1/2

-   Think constantly about reproducibility

    -   Your script files should run entirely without errors (start with your computer!)
    -   Rely on relative paths (no hard coding)
    -   RStudio projects help, use them!
    -   You may create a "package" folder to ensure compatibility across versions/platforms. Various options here.

-   Don't Repeat Yourself (DRY)

    -   Do not copy-paste pieces of code within the same script file (Solution: create a function and use loops)

    -   Do not copy-paste pieces of code across script files (Solution: create a script file containing functions, and load when necessary)

    -   Copy-pasting code is a red flag

-   Give objects and folders simple and intuitive names

    -   Good names are short and informative
    -   Limit use of upper case and weird characters (MyModified.GDP.var or log_gdp?)


# Guiding Principles 2/2

-   Annotate your code concisely but generously

    -   Better to err on the side of too many annotations
    -   Should help understand the purpose of any significant piece of your code
    -   Annotate single lines when code is relatively complex

-   Make your code look good (i.e. easy to read)!

    - Keep consistent indentation and spacing
    - Exploit RStudio capabilities to "fold" your code
    - Avoid deep nesting when possible
    - Let your code "breathe" (leave some blank spaces!)
    - Favor several lines with short expressions than a long line with multiple expressions
    

    
# General tips on the process of writing code

-   Before starting, you should first *visualize* the work flow...

    -   *"If you don't know where you're going, any road'll take you there"* (based on *Alice in Wonderland*, Lewis Carroll)
    -   Have a clear idea of repetitive tasks
    -   Think how you want your project and code to be structured when done

-   ... then start writing your code interactively

    -   Writing and running code in the console and script file editor
    -   But do not lose sight of the forest for the trees

-   Do not run a "big" task without testing on a small test dataset

-   Constantly re-check your code runs *entirely* (not just little pieces)

-   Leave "breadcrumbs" for easier debugging

```{r}
lapply(1:3, function(x1) {
  lapply(c("a","b"), function(x2) {
    print(paste(x1,"-",x2))
    # Meat of the code here
  })
})
```


# Getting started with your homework

1.  Read instructions in Canvas

2.  Create a project folder and subfolders

    -   ../hw1
    -   ../hw1/data

3.  Download data and place it in the data folder

4.  Create an RStudio project

5.  Create a new script file

6.  Visualize work flow and create sections

7.  Start coding (following guiding principles)!

